---
title: "Code structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Code structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette is aimed at developers who want to understand how the mizerSeasonal
package is implemented, contribute to it, or write their own extensions. It
describes the source layout, explains how the package hooks into mizer's
extension mechanism, and highlights the subtle or surprising aspects of the
implementation.

## Source layout

The package has seven R source files, listed here in the order they are collated
(as specified by the `Collate` field in `DESCRIPTION`):

| File | Contents |
|---|---|
| `mizerSeasonal-class.R` | S4 marker class definitions: `mizerSeasonal` and `mizerSeasonalSim` |
| `animate.R` | `animateGonadSpectra()` and two private helpers |
| `data.R` | Documentation for the `datta_params` dataset |
| `mizerSeasonal-package.R` | Package-level documentation, `@import` directives, and `.onLoad` hook |
| `plots.R` | `getTimeseries()`, `plotRDI()`, `plotRDD()`, `plotGonadsVsTime()` |
| `release_functions.R` | The four release-rate functions |
| `seasonal.R` | Core machinery: `setSeasonalReproduction()`, `gonadDynamics()`, `projectRDI.mizerSeasonal()`, `projectEncounter.mizerSeasonal()`, `seasonalBevertonHoltRDD()`, `seasonalVonMisesRDD()`, `seasonal_resource_semichemostat()`, and two private helpers |

`mizerSeasonal-class.R` is collated first so that the S4 class definitions are
available before any code that references them. `animate.R` uses unexported mizer
functions (`get_time_elements`) and dplyr verbs imported at the package level in
`mizerSeasonal-package.R`.

## How the package hooks into mizer

Mizer ships a formal extension API. The key functions used by this package are:

- **`registerExtension(pkgname, ...)`** — called once in `.onLoad` to register
  the package as a mizer extension for the current R session.
- **`setComponent(params, component, initial_value, dynamics_fun)`** — registers
  a new dynamical state variable (here, the gonadic mass matrix).
- **`setRateFunction(params, rate, fun)`** — replaces one of mizer's standard
  rate calculators with a custom one. Used here only for the RDD function;
  RDI and Encounter are handled via S3 dispatch instead (see below).
- **`coerceToExtensionClass(params)`** — promotes a `MizerParams` object to the
  registered S4 extension class (here, `mizerSeasonal`), enabling S3 method
  dispatch on `project*` generics.
- **`other_params(params)`** / **`other_params(params) <-`** — a named list for
  arbitrary configuration attached to a `MizerParams` object.

### The `.onLoad` hook

```{r}
.onLoad <- function(libname, pkgname) {
    mizer::registerExtension(pkgname, requirement = "sizespectrum/mizerSeasonal")
}
```

This runs when the package is loaded and tells mizer about the extension. It is
what makes `coerceToExtensionClass()` know which S4 class to promote to.

### S4 marker classes

Two S4 classes are defined in `mizerSeasonal-class.R`:

```{r}
setClass("mizerSeasonal",    contains = "MizerParams")
setClass("mizerSeasonalSim", contains = "MizerSim")
```

These are *marker* classes: they carry no additional slots. Their sole purpose
is to enable S3 method dispatch on the `project*` generics. When mizer calls
e.g. `projectRDI(params, ...)`, it dispatches to `projectRDI.mizerSeasonal`
because `params` has class `mizerSeasonal`. Similarly, `project()` returns a
`mizerSeasonalSim` because `MizerSim` objects are coerced to the registered
extension sim class.

### `setSeasonalReproduction()`

`setSeasonalReproduction()` wires everything together:

```{r}
setSeasonalReproduction <- function(params,
                                    release_func = "seasonalVonMisesRelease",
                                    RDD = "seasonalBevertonHoltRDD",
                                    include_gonads = TRUE) {
    initial <- initialN(params)
    initial[] <- 0

    other_params(params)$release_func  <- release_func   # <-- config
    other_params(params)$include_gonads <- include_gonads

    p <- setComponent(params,                            # <-- new state variable
                      component = "gonads",
                      initial_value = initial,
                      dynamics_fun = "gonadDynamics") |>
        setRateFunction("RDD", RDD)                      # <-- replace RDD only

    p@extensions <- mizer::getRegisteredExtensions()
    p <- mizer::coerceToExtensionClass(p)                # <-- promote to mizerSeasonal
    return(p)
}
```

Note that `setRateFunction("RDI", ...)` and `setRateFunction("Encounter", ...)`
are **not** called. Instead, those rates are intercepted via the S3 methods
`projectRDI.mizerSeasonal` and `projectEncounter.mizerSeasonal` described in the
next two sections. The RDD function is still set via `setRateFunction` because
`projectRDD.MizerParams` already dispatches internally through
`params@rates_funcs$RDD`; no S3 override is needed there.


### Where the release function name is stored

The choice of release function is stored as a character string inside
`other_params()`:

```{r}
other_params(params)$release_func  # e.g. "seasonalVonMisesRelease"
```

Internally mizer stores this in `params@other_params$other$release_func`, but
you should always use the `other_params()` accessor rather than the `@` slot
directly, because the slot layout may change across mizer versions.

At runtime, `gonadDynamics()` and `projectRDI.mizerSeasonal()` retrieve the function with:

```{r}
release_func <- get0(other_params(params)$release_func)
```

`get0()` is used instead of `get()` because it returns `NULL` rather than
throwing an error when the name is not found. If `release_func` is `NULL` the
code will fail on the next line, which gives a slightly clearer error than the
one that `get()` would raise.

### Where the initial gonadic mass lives

The initial gonad matrix (all zeros) is stored in:

```{r}
params@initial_n_other$gonads  # matrix (species x sizes)
```

This is mizer's standard storage for the initial state of `other` components.
It is set by `setComponent()` and is what `project()` uses to initialise the
simulation.

## The gonadic dynamics solver

`gonadDynamics()` solves the PDE for the per-capita gonadic mass $q(w,t)$ at
each time step. The derivation of the PDE is given in the `gonadic_mass_dynamics`
vignette; the result is a tridiagonal linear system of the form

$$a_{ij}\, q_{j-1} + b_{ij}\, q_j = s_{ij}$$

where the coefficients are:

```{r}
# a_{ij} = -g_i(w_j) * dt / dw_j       (sub-diagonal: growth advection)
a <- sweep(-rates$e_growth * dt, 2, params@dw, "/")

# b_{ij} = 1 + g_i(w_j) * dt / dw_j + r_i * dt   (diagonal: advection + release)
b <- 1 + sweep(rates$e_growth * dt, 2, params@dw, "/") + r * dt

# s_{ij} = q_{ij} + e_repro_{ij} * dt   (right-hand side: old state + investment)
s <- n_other$gonads + rates$e_repro * dt
```

The system is then solved by calling mizer's internal C++ routine:

```{r}
mizer:::inner_project_loop(no_sp = no_sp, no_w = no_w,
                            n = n_other$gonads,
                            A = a, B = b, S = s,
                            w_min_idx = params@w_min_idx)
```

The `:::` call is intentional: `inner_project_loop` is the same forward-sweep
solver that mizer uses for the fish number density, and there is no public
wrapper for it. This creates a hard dependency on mizer's internal layout, so
any future mizer refactor that renames or moves this function will break
`gonadDynamics`. The dependency is accepted because (a) the physics of the two
PDEs is identical, so reusing the solver is correct, and (b) the alternative
would be to duplicate several hundred lines of compiled C++.

## Release functions

All four release functions share the same signature:

```{r}
release_func <- function(t, params, ...)
```

and return a numeric vector of length equal to the number of species, giving the
mass-specific gonad release rate $r_i(t)$ for each species $i$ at time $t$.

### Periodicity mechanisms

All release functions produce a rate that is periodic with a period of exactly
one year. They achieve this in slightly different ways:

| Function | Periodicity mechanism |
|---|---|
| `seasonalVonMisesRelease` | $\cos(2\pi(t-\mu))$ is inherently $1$-periodic |
| `seasonalBetaHazardRelease` | `t - floor(t)` maps any $t$ to $[0,1)$ |
| `seasonalBetaRelease` | `t - floor(t)` |
| `seasonalGaussianRelease` | `t - trunc(t)` |

`trunc()` and `floor()` agree for positive arguments (both truncate towards
zero), so the last two are equivalent in practice. The inconsistency between
them is a historical artefact rather than a deliberate design choice.

### The von Mises rate at the seasonal peak and trough

The von Mises release rate is

$$r(t) = r_0 \frac{e^{\kappa \cos(2\pi(t-\mu))}}{2\pi I_0(\kappa)}$$

The cosine reaches its maximum of $+1$ at $t = \mu$ and its minimum of $-1$ at
$t = \mu + 0.5$, so:

$$r(\mu) = \frac{r_0\, e^{\kappa}}{2\pi I_0(\kappa)}, \qquad
  r(\mu + 0.5) = \frac{r_0\, e^{-\kappa}}{2\pi I_0(\kappa)}$$

The ratio of peak to trough rate is $e^{2\kappa}$, so a concentration of
$\kappa = 2$ gives a ratio of about $54$.

## Density-dependent reproduction: `seasonalBevertonHoltRDD`

This function is a time-varying version of mizer's built-in `BevertonHoltRDD`.
It applies the formula

$$R_{dd}(t) = \frac{R_{di}(t)\, R_{\max}(t)}{R_{di}(t) + R_{\max}(t)}$$

where $R_{\max}(t)$ is looked up from a 2-D array stored in
`other_params(params)$r_max`. That array must have a `time` dimension in its
`dimnames`:

```{r}
# Required structure of other_params(params)$r_max:
times <- seq(0, 0.9, by = 0.1)
r_max <- matrix(
    values,
    nrow = length(times), ncol = no_sp,
    dimnames = list(time = as.character(round(times, 5)),
                    species = sp_names)
)
other_params(params)$r_max <- r_max
```

The time lookup converts the fractional year to a character key:

```{r}
t_name <- as.character(round(t - floor(t), 5))
r_max  <- other_params(params)$r_max[t_name, ]
```

Two practical consequences follow from this design:

1. The `r_max` array must be pre-populated for every time step at which the
   simulation will be evaluated (i.e., every multiple of `dt` within `[0, 1)`).
   Missing entries cause an explicit `stop()`.
2. Floating-point rounding in `as.character(round(x, 5))` can produce
   surprising keys. For example, `as.character(round(0.1, 5))` is `"0.1"` on
   most platforms, but very small floating-point errors in the simulation clock
   could push `t - floor(t)` to something like `0.09999999999` which rounds to
   `"0.1"`. This is typically safe, but worth being aware of if `dt` is very
   small (e.g., `dt = 0.001`).

## `seasonalVonMisesRDD`: an unusual signature

This function calculates a von Mises-distributed reproduction rate that is
**independent of the actual egg production**. Its intended use is as a
placeholder during model calibration (before a proper $R_{\max}$ has been
estimated). Its signature is:

```{r}
seasonalVonMisesRDD <- function(params, t, ...)
```

Mizer calls all RDD functions as:

```{r}
do.call(params@rates_funcs$RDD,
        list(rdi = rdi, params = params, t = t, ...))
```

Because `rdi` is not in `seasonalVonMisesRDD`'s formal argument list, it is
silently absorbed by `...` and ignored. This works, but means that:

- The function cannot be used for a calibrated simulation: it ignores egg
  production entirely and can produce densities that are inconsistent with the
  gonadic dynamics.
- The documentation explicitly warns that it "should be replaced by another
  reproduction rate function before simulating the dynamics."

## `projectRDI.mizerSeasonal`: gonadic-release egg production

This S3 method replaces mizer's standard energy-based RDI calculation entirely.
It does **not** call `NextMethod()`. Instead it computes egg production directly
from the gonadic release:

```{r}
projectRDI.mizerSeasonal <- function(params, n, n_pp, n_other, t = 0,
                                     e_growth, mort, e_repro, ...) {
    release_func <- get0(other_params(params)$release_func)
    r <- release_func(t, params)            # mass-specific release rate (species vector)
    total <- drop((sweep(n_other$gonads, 1, r, "*") * n) %*% params@dw)
    0.5 * (total * params@species_params$erepro) /
        params@w[params@w_min_idx]
}
```

` r * gonads * n * dw` integrates to give the total gonadic mass released per
unit time per species. Dividing by the egg weight and multiplying by `erepro /
2` (the female fraction × reproductive efficiency) converts mass to egg numbers.
The `e_growth`, `mort`, and `e_repro` arguments are present only for generic
compatibility and are not used.

## `projectEncounter.mizerSeasonal`: prey mass includes gonads

This S3 method is **additive**: it calls `NextMethod()` to get the standard
encounter rate and then appends the extra encounter arising from the gonadic mass
carried by prey fish. The approach is mathematically exact because the encounter
integral is linear in prey biomass:

$$E(w_\text{somatic} + q) = E(w_\text{somatic}) + E(q)$$

so the two contributions can be computed separately and summed.

The gonad contribution is only added when `include_gonads = TRUE` (set in
`setSeasonalReproduction`). When it is `FALSE`, the method returns `NextMethod()`
unchanged.

Note that the gonad contribution is zero if the predator–prey interaction matrix
entry for that predator–prey species pair is zero. In a single-species model
built with `newSingleSpeciesParams()` the interaction matrix has `interaction[1,1] = 0`
by default (no self-predation), so the gonad contribution is exactly zero and
`projectEncounter.mizerSeasonal` returns the same result as `mizerEncounter`. To
observe a difference you need at least one nonzero fish-on-fish interaction.

The method contains two code paths — one using the FFT convolution (the fast
default) and one using direct matrix multiplication (used when the user has
supplied a custom `pred_kernel`). In the FFT path, `prey_gonad` is swept by
`dw_full` only (not `w_full * dw_full`) because `gonads` already represents
absolute mass rather than the mass density $w \cdot N(w)$ used for somatic
biomass.

## `seasonal_resource_semichemostat`: configuration via `resource_params$rp`

This function modulates the resource carrying capacity with a von Mises seasonal
signal. It reads its parameters from a named sub-list `rp` of the
`resource_params` slot:

```{r}
params@resource_params$rp <- list(kappa = 2, mu = 0.25, maxR = 0.5)
```

| Key | Meaning |
|---|---|
| `kappa` | Concentration of the von Mises peak (higher = sharper season) |
| `mu` | Time of peak resource abundance within the year |
| `maxR` | Fractional amplitude: capacity is multiplied by $(1 + \text{maxR} \cdot \text{vonMises})$ |

This differs from how the release functions receive their parameters (via
`species_params` columns). The choice of `resource_params$rp` is because the
resource has no species dimension, so `species_params` is not appropriate.

## Data extraction: `getTimeseries`

`getTimeseries(sim, func)` loops over the saved time steps of a `MizerSim`
object and calls `func` at each step, reconstructing the per-time-step state
vectors that mizer does not store directly. The reconstruction follows the same
pattern that mizer uses internally:

```{r}
for (i in time_indices) {
    n <- sim@n[i, , ]
    dim(n) <- dim(params@initial_n)          # restore matrix shape
    n_other <- sim@n_other[i, ]              # named list of other components
    names(n_other) <- dimnames(sim@n_other)[[2]]
    value[i, ] <- func(params,
                       n = n,
                       n_pp = sim@n_pp[i, ],
                       n_other = n_other,
                       t = times[i])
}
```

The `func` argument defaults to `mizer::getRDI`, so `getTimeseries(sim)` returns
the RDI time series. Pass `mizer::getRDD` to get the RDD time series, which is
what `plotRDD` does.

## Plotting functions

`plotRDI` and `plotRDD` delegate the data extraction to `getTimeseries` and then
rely on mizer's internal `plotDataFrame` helper for the actual rendering. The
data frame passed to `plotDataFrame` must have its columns in the specific order
`(Year, Value, Species)`, which is why both functions do an explicit column
reorder:

```{r}
plot_dat <- plot_dat[, c(1, 3, 2)]  # reorder to (Year, Value, Species)
```

`plotGonadsVsTime` is a simpler base-graphics function that plots the per-capita
gonadic mass at the species' maximum body size through time. When `time_range` is
supplied as a two-element vector it is interpreted as `(t_min, t_max)` and
passed to mizer's `get_time_elements()`, which returns a logical index vector
over all saved times — not a count. The number of selected time steps is
therefore `sum(time_elements)`, not `length(time_elements)`, and the code must
use `length(qf)` (the number of selected gonad matrices) as the row dimension of
the output matrix:

```{r}
qf     <- sim@n_other[time_elements, "gonads"]   # list, length = sum(time_elements)
qs_mat <- matrix(..., nrow = length(qf), ...)    # NOT length(time_elements)
```

## Test suite

The tests live in `tests/testthat/` and are organised as follows:

| File | What it tests |
|---|---|
| `helper-setup.R` | Shared fixtures: `base_params`, `seasonal_params`, `seasonal_sim` |
| `test-extension.R` | Extension mechanism: class promotion, S3 dispatch, `project()` integration |
| `test-release_functions.R` | Mathematical properties of all four release functions |
| `test-seasonal.R` | `setSeasonalReproduction`, `gonadDynamics`, `projectRDI.mizerSeasonal`, `seasonalBevertonHoltRDD`, `seasonalVonMisesRDD`, `projectEncounter.mizerSeasonal`, `seasonal_resource_semichemostat` |
| `test-plots.R` | `getTimeseries`, `plotRDI`, `plotRDD`, `plotGonadsVsTime`, `animateGonadSpectra` |

`test-extension.R` uses its own `local_params()` helper (built on
`NS_species_params`) rather than the shared `seasonal_params` fixture. This keeps
the extension-mechanism tests self-contained and avoids depending on the
single-species model that the shared fixture uses.

The fixture `seasonal_params` uses `seasonalVonMisesRDD` as the RDD function
(rather than the default `seasonalBevertonHoltRDD`) to avoid having to
construct the `r_max` time-series array that `seasonalBevertonHoltRDD` requires.
Tests of `seasonalBevertonHoltRDD` itself construct a minimal `r_max` inline.

The fixture `seasonal_sim` is produced by
`project(seasonal_params, t_max = 2, dt = 0.1, t_save = 0.5)`, giving five
saved time points (0, 0.5, 1, 1.5, 2). Keeping `t_max` small and `t_save`
coarse makes the helper fast to load without sacrificing coverage of the plot
functions that need a `MizerSim` object.
