---
title: "mizerSeasonal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mizerSeasonal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mizerSeasonal)
```

To demonstrate the use of the package we first set up a standard trait-based model, turn off density dependence in reproduction and run to steady state.
```{r}
p <- newTraitParams()
# Turn of density dependence for now
p <- setRateFunction(p, "RDD", "noRDD")
# Find steady state
p <- steady(p)
```

We will set up seasonal reproduction with the default gaussian mass-specific reproduction rate
$$
r(w, t) = r_0 \exp{\left(-\frac{(\lfloor{t}\rfloor-t_0)^2}{2\sigma^2}\right)}
$$
Because mizer measures time in years, $\lfloor t\rfloor$ gives the time within the year.

We need to specify the new species parameters $r_0, \sigma$ and $t_0$ (which are denoted by `sr_r0`, `sr_sigma` and `sr_t0` respectively) and then call `setSeasonalReproduction()`.
```{r}
p@species_params$sr_r0 <- 5
p@species_params$sr_sigma <- sqrt(1/2000)
p@species_params$sr_t0 <- 0.2
p <- setSeasonalReproduction(p)
```

The resulting MizerParams object can be projected into the future as usual with `project()` to produce a MizerSim object.
```{r}
sim <- project(p, t_max = 3, t_save = 0.05, dt = 0.01)
```

Let us take a look at the seasonal reproduction rate $R_p(t)$
```{r}
plotRDI(sim, log = FALSE)
```

When we animate the simulation result we see the effect travelling up the size spectrum
```{r}
animateSpectra(sim, power = 2)
```

On the logarithmic scale the change in the gonadic mass is not so easy to see
```{r}
animateGonadSpectra(sim)
```

